Blocks of extended ancestry around selected loci in hybrid zones

### 2016-02-22 drafts/hybrid_zone.tex
minor fixes to methods/results

Similar to the model presented by \cite{Sedghifar2015}, we consider a lineage sampled in the present day, and determine its ancestry by its position in space at the time of first contact, $\tau$ generations in the past (Figure~\ref{Fig:schematic}). 
The ancestry at locus $L$ is determined by its state, so that a lineage with identity $L_A$ is of ancestry $A$ and a lineage with identity $L_B$ is of ancestry $B$ regardless of its sampling location (we assume that at time $\tau$ the two species are fixed for alternate alleles at this locus). 
For every other locus, we keep track of $z$, the identity at $L$, across time and space. If at time $\tau$ a locus is on the same background as $L_A$, it is considered to be of ancestry $A$, and of ancestry $B$ otherwise. Recombination between two loci is modeled as a poisson process with rate $r$. The identity $z$ for locus $K$ changes when a recombination event occurs between $L$ and $K$ in an individual that is heterozygous at locus $L$.

@@

\paragraph{Establishment of  \alisa{a cline at the selected locus}}
Our simple model of underdominance considers a single locus, with alternate alleles fixed in the two parental populations,
for which heterozygotes have fitness $1-s$, and homozygotes are of equal fitness.
The deterministic theory predicts that after secondary contact,
these alleles will form a stable cline \citep{Barton1979}.
To study gene flow past this cline we \alisa{turn our attention to} how the cline itself is formed.
Let $p(x,t)$ denote the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, \alisa{(Will switching between backwards and forwards in time not get confusing?)}
and that $p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
(In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
Let $\sigma^2$ be the mean squared distance between parent and offspring \alisa{(i.e. the variance of the distribution of dispersal distances)}.
Assuming that: (i) alleles locally assort into diploids randomly, (ii) habitat and dispersal are homogeneous, and (iii) $s$ and $\sigma$ are small, then the commonly-used equation that $p$ solves is

\begin{align} \label{eqn:cline_pde}
    \dot p = \frac{\sigma^2}{2} \Delta p + s p (1-p) (2p-1) ,
\end{align}
where $\dot p(t,x)$ is the time derivative and $\Delta p$ is the Laplacian,
as derived in \citep{Bazykin1969}; for discussion of diploidy see \citep{diploidcline} \alisa{not sure which reference this is}.
Although there is not an exact analytic solution to this equation, a good approximation to the equilibrium solution is that
$\lim_{t \to \infty} p(x,t) \approx (1+\tanh(-2x\sqrt{s}/\sigma))/2$ \citep{Bazykin1969}. This relies on numerous approximations, but importantly, the stable cline has width of order $\sigma/\sqrt{s}$,
and decays exponentially. Note that rescaling space and time by $\sigma/s$ and $1/s$ respectively in equation \eqref{eqn:cline_pde} doesn't affect this solution, and so the cline establishes over a timescale of the order $1/s$. In other words, the cline is established through diffusion, which is slowed down somewhat by selection against heterozygotes. \alisa{I don't think I understand the independence from $\sigma$?}

@@

From previous results we expect that far away from the zone center, variants under negative selection will be rare \citep[for demonstration of this theoretical result, see e.g.][]{May1975,Slatkin1973,Barton??}, and correspondingly, the width of clines at closely linked sites are impacted (Fig:~\ref{alleleFreq_tau100}).

@@

A linked cline will begin to flatten out if there is an appreciable chance that a lineage at the linked locus initially linked to an $A$ allele 
at some point recombined from an individual carrying the $B$ allele.
This implies that loci closer than $1/\tau$ to the selected site won't have a substantially wider cline than the selected cline.
Of course, this calculation assumes that every generation provides an opportunity for recombination on the other background, but the only way to move linked lineages between selected backgrounds is via recombination in a heterozygote.

@@

In the context of our model, we expect that\alisa{, in the absence of selection,} $l_A$ across the genome will tend to be shorter the further one into the $B$ side of the cline,
because they have had more opportunities to recombine with $B$ haplotypes. 
However, stretches of $A$ ancestry that contain a selected site are expected to be longer
than those that do not at the same spatial location,
because lineages containing the selected site must have been inherited from the $A$ side of the cline recently --
concretely, they move at speed roughly $\sigma/\sqrt{s}$,
so we expect (selected) $A$ alleles at distance $x$ from the cline center
to have last had an ancestor on the $A$ side of the cline around $x \sqrt{s}/\sigma$ generations ago \alisa{(compared to $\sigma$?? for a neutral allele)},
and so the enclosing $A$ haplotype is no longer than $\sigma/(x \sqrt{s})$.
\plr{(check predicted decay of mean length $1/x$ in simulations).}
For linked haplotypes not containing the selected site, the situation is more complicated, and harder to intuit.

@@

\subsection*{The number of `unfit' ancestors}
Although the theoretical model presented here does not account for coalescence, we are able to partially trace the genealogy of haplotypes in our simulations. In particular, we wish to know the relative number of ancestors from $\tau$ generations ago that are represented in present day populations at a given locus, given the frequency of a particular ancestry. This is a reflection of the average size of a haplotype family. We find that, as distance from the zone center increases, the average family size of an ancestor from time $\tau$ decreases (Figure~\ref{Fig:family_size}). One interpretation of this observation is that migrants tend to leave few offspring before being removed from the population by selection. This can potentially provide additional signal for characterizing lineages strongly influenced by selection.

>>>>

Similar to our recent model of a neutral hybrid zone \cite{Sedghifar2015}, we consider a lineage sampled in the present day, and determine its ancestry by its position in space at the time of first contact, $\tau$ generations in the past (Figure~\ref{Fig:schematic}). 
The ancestry at locus $L$ is determined by its state, so that a lineage with identity $L_A$ is of ancestry $A$ and a lineage with identity $L_B$ is of ancestry $B$ regardless of its sampling location (we assume that at time $\tau$ the two species are fixed for alternate alleles at this locus). 
For every other locus, we keep track of $z$, the identity at $L$, across time and space. If at time $\tau$ a locus is on the same background as $L_A$, it is considered to be of ancestry $A$, and of ancestry $B$ otherwise. Recombination between two loci is modeled as a poisson process with rate $r$. The identity $z$ for locus $K$ changes when a recombination event occurs between $L$ and $K$ in an individual that is heterozygous at locus $L$.

@@

\paragraph{Establishment of  \alisa{a cline at the selected locus}}
Our simple model of underdominance considers a single locus, with alternate alleles fixed in the two parental populations,
for which heterozygotes have fitness $1-s$, and homozygotes are of equal fitness.
The deterministic theory predicts that after secondary contact,
these alleles will form a stable cline \citep{Barton1979}.
To study gene flow past this cline, we \alisa{turn our attention to} how the cline itself is formed.
Let $p(x,t)$ denote the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, \alisa{(Will switching between backwards and forwards in time not get confusing?)}
and that $p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
(In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
Let $\sigma^2$ be the mean squared distance between parent and offspring \alisa{(i.e. the variance of the distribution of dispersal distances)}.
Assuming that: (i) alleles locally assort into diploids randomly, (ii) habitat and dispersal are homogeneous, and (iii) $s$ and $\sigma$ are small, then the commonly-used equation that $p$ solves is

\begin{align} \label{eqn:cline_pde}
    \dot p = \frac{\sigma^2}{2} \Delta p + s p (1-p) (2p-1) ,
\end{align}
where $\dot p(t,x)$ is the time derivative and $\Delta p$ is the Laplacian,
as derived in \citep{Bazykin1969}; for discussion of diploidy see \citep{diploidcline} \alisa{not sure which reference this is}.
Although there is not an exact analytic solution to this equation, a good approximation to the equilibrium solution is that
$\lim_{t \to \infty} p(x,t) \approx (1+\tanh(-2x\sqrt{s}/\sigma))/2$ \citep{Bazykin1969}. This relies on numerous approximations, but importantly, the stable cline has width of order $\sigma/\sqrt{s}$,
and decays exponentially. Note that rescaling space and time by $\sigma/s$ and $1/s$ respectively in equation \eqref{eqn:cline_pde} does not affect this solution, and so the cline establishes over a timescale of the order $1/s$. In other words, the cline is established through diffusion, which is slowed down somewhat by selection against heterozygotes. \alisa{I don't think I understand the independence from $\sigma$?}

@@

From previous results, we expect that far away from the zone center, variants under negative selection will be rare \citep[for demonstration of this theoretical result, see e.g.][]{May1975,Slatkin1973,Barton??}, and correspondingly, the width of clines at closely linked sites are impacted (Fig:~\ref{alleleFreq_tau100}).

@@

A linked cline will begin to flatten out if there is an appreciable chance that a lineage at the linked locus initially coupled with an $A$ allele 
at some point recombined from an individual carrying the $B$ allele.
This implies that loci closer than $1/\tau$ to the selected site will not have a substantially wider cline than the selected cline.
Of course, this calculation assumes that every generation provides an opportunity for recombination on the other background, but the only way to move linked lineages between selected backgrounds is via recombination in a heterozygote.

@@

In the context of our model, we expect that\alisa{, in the absence of selection,} $l_A$ across the genome will tend to be shorter the further one into the $B$ side of the cline,
because they have had more opportunities to recombine with $B$ haplotypes. 
However, we expect that stretches of $A$ ancestry containing a selected site will be longer
than those that do not at the same spatial location,
because lineages containing the selected site must have been inherited from the $A$ side of the cline recently. 
Concretely, these lineages move at speed roughly $\sigma/\sqrt{s}$,
so we expect (selected) $A$ alleles at distance $x$ from the cline center
to have last had an ancestor on the $A$ side of the cline around $x \sqrt{s}/\sigma$ generations ago \alisa{(compared to $\sigma$?? for a neutral allele)},
and so the enclosing $A$ haplotype is no longer than $\sigma/(x \sqrt{s})$.
\plr{(check predicted decay of mean length $1/x$ in simulations).}
For linked haplotypes not containing the selected site, the situation is more complicated, and harder to intuit.

@@

\subsection*{The number of `unfit' ancestors}
Although our analytical model does not account for coalescence, we are able to partially trace the genealogy of haplotypes in our simulations. In particular, we wish to know the relative number of ancestors from $\tau$ generations ago that are represented in present day populations at a given locus, given the frequency of a particular ancestry. This is a reflection of the average size of a haplotype family. We find that, as distance from the zone center increases, the average family size of an ancestor from time $\tau$ decreases (Figure~\ref{Fig:family_size}). One interpretation of this observation is that migrants tend to leave few offspring before being removed from the population by selection. This can potentially provide additional signal for characterizing lineages strongly influenced by selection.
